---
title: "Assignment-1"
author: "Aniket Guha Roy-19200164"
date: "10/11/2019"
output:
  pdf_document: default
  html_document: default
---
##  Manipulation  
1. Load the dataset EurostatCrime2017.csv. Notice that the ﬁrst column of the csv ﬁle contains the names of the countries that must be read as row names [Hint: Load in the ﬁle using the function read.csv]
```{r}
setwd("C:/Users/anike/Desktop/R module/Assignments/Assignment")
EuroCrime = read.csv("EurostatCrime2017.csv",header = T, row.names = 1)

```
2. What is the size (number of rows and columns) and the structure of this dataset?  
number of rows: 41
number of columns : 11
```{r}
dim(EuroCrime)
str(EuroCrime)
```

3. Produce appropriate commands to answer the following questions:
(i) For some countries Theft includes also burglary, and theft of motorised land vehicle, in others they are recorded separately. Add a new column called All Theft which contains the sum of all the crimes that have a theft component: • Theft, • Theft of a motorized land vehicle, • Burglary, • Burglary of private residential premises Please consider NA values as 0 in this case. [Hint: you may want to use the function apply or rowSums, check their help ﬁles to see how to deal with missing values.]
```{r}
EuroCrime$AllTheft = rowSums(EuroCrime[,c("Theft","Theft_of_a_motorized_land_vehicle",
                      "Burglary","Burglary_of_private_residential_premises")],na.rm = T)
```

(ii)Remove the columns: Theft, Theft of a motorized land vehicle, Burglary, and Burglary of private residential premises
```{r fig.height = 5, fig.width = 4, fig.align = "center"}
EuroCrime[ ,c("Theft","Theft_of_a_motorized_land_vehicle","Burglary",
              "Burglary_of_private_residential_premises")]=NULL
head(EuroCrime)
```

4. Work with the dataset you created in question 3ii, and list the countries that contain any missing data
Ans: Denmark"                "France"                 "Croatia"                "Hungary"               
  "Netherlands"            "Austria"                "Poland"                 "Portugal"              
  "Sweden"                 "England_and_Wales"      "Iceland"                "Liechtenstein"         
 "Norway
```{r}
  row.names(EuroCrime[!complete.cases(EuroCrime),])

```

5. Remove the countries with missing data from the dataframe
```{r}
EuroCrime=EuroCrime[complete.cases(EuroCrime),]
```

6. How many observations and variables are in this new dataframe  
number of rows: 25
number of columns:8
```{r  fig.height = 6, fig.width = 6, fig.align = "center"}
dim(EuroCrime)
```

## Analysis  
1.According to these data what were the 3 most common crimes in Ireland in 2017?  
Ans: 
1:All_Theft
2.Unlawful Acts
3.Assault

```{r}
sort(EuroCrime[c("Ireland"),],decreasing = T)[1:3]
```

2.Which country has the highest overall record of oﬀences (per hundred thousand inhabitants)?
Ans: Switzerland
```{r}
rownames(EuroCrime)[rowSums(EuroCrime) == max(rowSums(EuroCrime))]
```


## Creativity  
1.First we are going to take a look at the top countries having highest overall record of offences:  
We already knew Switzerland was the country with higest offenses recorded but we get an idea of other top countries and how they match up to SWitzerland as well 
```{r fig.height = 6, fig.width = 8, fig.align = "center"}
barplot(sort(rowSums(EuroCrime),decreasing = T)[1:10],ylab = "Frequency", 
        main = "Top 10 Countries with most crimes",las=2, col = "lightblue")
```

2. Rank of the Crimes  
This graph gives us an idea of which is the most common committed crime, i.e. Theft/Burglary. Next most common crime is Unlawful_acts_involving_controlled_drugs_or_precursors followed by Assault.But Unlawful acts is quite less than Theft. Hence overall if we could bring down Thefts, the crime rate might get reduced overall.

```{r fig2, fig.height = 6, fig.width = 8, fig.align = "center"}
barplot(colSums(EuroCrime), main = "Overall Rank of the Crimes", 
        ylab = "Frequency",las=2, cex.names = 0.5,col = "lightblue")
```

3. All_Thefts  
This graph gives an idea of the most committed crime-All Thefts  
- We infer that there are no outliers.
- All the observations lie between 1st and 3rd quadrants 
- Skewed right or positively skewed
- Median slightly above 1000  
INFERENCE: 
- There are no countries in the data which have a frequency of Theft less than approx 600 and more than roughly 2100.
- Hence we get to know most of the countries in the data have an occurence of thefts/ burglary more than 1000 times. 
```{r}
boxplot(EuroCrime$AllTheft, main = "ALL_Thefts", ylab = "frequency")
```

