---
title: "Predictive Analytics Assignment"
author: "Aniket Guha Roy -19200164"
date: "11/17/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}

install.packages("psych")
install.packages("car")
install.packages("tidyverse")
install.packages("olsrr")
install.packages("GGally")
install.packages("lmtest")
install.packages("corrplot")
library(corrplot)
library(lmtest)
library(car)
library(psych)
library(ggplot2)
library(olsrr)
library(GGally)
```



```{r house}
house = read.csv(file.choose(),header = T)
str(house)
house$Price = house$ï..Price
house$ï..Price = NULL
```

## Exploratory Data Analysis:
1.  Using a boxplot, histogram and summary. Describe the distribution of the sales price of the houses.

```{r pressure, echo=FALSE}
summary(house$Price)

par(mfrow=c(1,2))
boxplot(house$Price,xlab = "Price of houses")
hist(house$Price,breaks = "FD",freq = F, xlab = "House Price",main = "Histogram of house price")
lines(density(house$Price), lwd=2, col="blue")


```
2.Convert all the categorical variables to factors. Using the summary and a boxplot describe how sales prices vary with respect to the number of bedrooms, bathrooms, garage size and school.

```{r }
house[,c("Bath","Bed","Garage")] = lapply(house[,c("Bath","Bed","Garage")],as.factor)
str(house)

summary(house$bath)

by(house$Price,house$Bath,summary)

by(house$Price,house$Bath,summary)

by(house$Price,house$Bath,summary)

by(house$Price,house$Bath,summary)

boxplot(house$Price~house$Bath)
boxplot(house$Price~house$Bed)
boxplot(house$Price~house$Garage)
boxplot(house$Price~house$School)


```

3. Using the summary, correlation and the pairs plots discuss the relationship between the response sales price and each of the numeric predictor variables

```{r}
summary(house$Size)
summary(house$Lot)
summary(house$Year)
pair_plot = ggpairs(house[,c("Price","Size","Lot","Year")],columnLabels = c("Price","Size","Lot","Year"))
show(pair_plot)
cor.test(house$Price,house$Size, method="pearson")
cor.test(house$Price,house$Lot, method="pearson")
cor.test(house$Price,house$Year, method="pearson")
```

##Regression Model
```{r}
fit<-lm(Price~.,data=house)
formula(fit)
summary(fit)
plot(fit)
```

##ANOVA

```{r}
anova(fit)
fit2=lm(Price ~ Size + Lot + Bath + Bed + Garage + School, data =house)
anova(fit,fit2)

```

##Diagnostics:
1.Check the linearity assumption by interpreting the added variable plots and component-plus-residual plots.
```{r}
avPlots(fit)
crPlots(fit)
```

2. Check the random/i.i.d. sample assumption by carefully reading the data description and computing the Durbin Watson test.
```{r}
dwtest(fit)
```

3.Check the collinearity assumption by interpreting the correlation and variance inﬂation factors.
```{r}
cor<-cor(house[,c("Size","Lot","Year")])
corrplot.mixed(cor)
vif(fit)
```

4. Check the zero conditional mean and homoscedasticity assumption by interpreting the studentized residuals vrs ﬁtted values plots and the studentized residuals vrs predictor variable plots.
```{r}
rstudent_res = rstudent(fit)
plot(fit$fitted.values,rstudent_res)
plot(house$Size,rstudent_res)
plot(house$Lot,rstudent_res)
plot(house$Year,rstudent_res)

```

5. Check the Normality assumption by interpreting the histogram and quantilequantile plot of the studentized residuals
```{r}
hist(rstudent_res)
qqnorm(rstudent_res)
qqline(rstudent_res)
```

##Leverage, Inﬂuence and Outliers

1. Use the leverage values and the leverage plots to see if there is any leverage points
```{r}
leverage_points = as.numeric(which(hatvalues(fit)>((2*7)/length(house$Price))))
leverage_points
leveragePlots(fit)
```

2. Use the inﬂuence plot to see if there is any inﬂuence points.

```{r}
influencePlot(fit, id.method="identify", main="Influence Plot")

```

3. Use the outlier test and outlier and leverage diagnostics plot to see if there is any outliers. Deal with the outliers if any are identiﬁed.

```{r}
outlierTest(fit)
ols_plot_resid_lev(fit)
ols_plot_cooksd_bar(fit)
```

##Expected Value, CI and PI:

1. Plot the observed house prices, their expected vale (ﬁtted value), conﬁdence intervals (in red) and prediction intervals (in blue). 
```{r}
c<-predict(fit,level = 0.95,interval = "confidence")
p<-predict(fit,level = 0.95,interval = "prediction")
ggplot(house,aes(y=house$Price,x=fitted(fit)))+geom_point()+
 stat_smooth(aes(y=c[,"upr"]),method = lm,se=F,col='red')+
 stat_smooth(aes(y=c[,"lwr"]),method = lm,se=F,col='red')+
 stat_smooth(aes(y=p[,"upr"]),method = lm,se=F,col='blue')+
 stat_smooth(aes(y=p[,"lwr"]),method = lm,se=F,col='blue')+
 geom_line(aes(y=c[,"fit"]),col='green')

```

