---
title: "R Assignment 2"
author: "Aniket Guha Roy"
date: "08/11/2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---
```{r library
}
install.packages("tidyverse")
library(ggplot2)
install.packages("ggpubr")
library(ggpubr)
```



2.Load in the data as an object called DublinAirport. Notice that you have to skip the
first 19 lines contained in the file before beginning to read data. Line 20 contains the
column names. Display the structure of the dataset

```{r}
setwd("C:/Users/anike/Desktop/UCD/R module/Assignments/Assignment/Assignment-2")
DublinAirport = read.csv("mly532.csv",header = T, skip = 19)
str(DublinAirport)
```

3.Transform the column months to a factor with labels given by the month names. [Two
hints: (1) look at the arguments levels and labels of the function factor. (2) you
may want to make use of the built-in constant month.name]

```{r}
DublinAirport$month <- factor(DublinAirport$month,levels = c(1,2,3,4,5,6,7,8,9,10,11,12),labels = month.name)
```

4.Use the aggregate function to compute which month has on average the highest and
the lowest Precipitation Amount.

```{r}
res = aggregate(rain~month,aa,sum)
res[order(res$rain),]
```

5.Create a new column which contains a factor indicating the season:
• Winter: December, January, February,
• Spring: March, April, May,
• Summer: June, July, August,
• Autumn: September, October, November
[Hint: again, look at the arguments levels and labels of the function factor.]
```{r}
DublinAirport$season <- factor(aa$month, levels =month.name[c(12,1,2,3,4,5,6,7,8,9,10,11)],labels = c("Winter","Winter","Winter","Spring","Spring","Spring","Summer","Summer","Summer","Autumn","Autumn","Autumn"))

```

6. Assign to the DublinAiport object the classes WeatherData and data.frame.

```{r}
class(DublinAirport) = c("WeatherData","data.frame")
```

7. Write an S3 summary method for an object of class WeatherData which produces
the following statistical summaries the rain, maxtp, mintp, maxgt variables split by
season: mean, standard deviation, minimum, maximum. Ignore the missing values in
the calculations.
Test your function on the DublinAirport data set and comment your findings

```{r}
summary.WeatherData = function(obj){
  Mean_statistic = aggregate(cbind(obj$rain,obj$maxtp,obj$mintp,obj$maxgt), by =list(obj$season),FUN = mean, na.rm = T)
  SD_statistic = aggregate(cbind(obj$rain,obj$maxtp,obj$mintp,obj$maxgt), by = list(obj$season), FUN = sd, na.rm = T)
  Minimum_statistic = aggregate(cbind(obj$rain,obj$maxtp,obj$mintp,obj$maxgt), by = list(obj$season), FUN = min, na.rm = T)
  Maximum_statistic = aggregate(cbind(obj$rain,obj$maxtp,obj$mintp,obj$maxgt), by = list(obj$season), FUN = max, na.rm = T)
  colnames(Mean_statistic) = c("Seasons","Rain_mean","Maxtp_mean","Mintp_mean","Maxgt_mean")
  colnames(SD_statistic) = c("Seasons","Rain_SD","Maxtp_SD","Mintp_SD","Maxgt_SD")
  colnames(Minimum_statistic) = c("Seasons","Rain_min","Maxtp_min","Mintp_min","Maxgt_min")
  colnames(Maximum_statistic) = c("Seasons","Rain_max","Maxtp_max","Mintp_max","Maxgt_max")
  print(Mean_statistic)
  print(SD_statistic)
  print(Minimum_statistic)
  print(Maximum_statistic)
}
summary.WeatherData(DublinAirport)
```


8. Create an S3 plot method for the class WeatherData that produces the following plots.
(1) Plot of the monthly Air Temperature (C) (maxtp, mintp).
(2) Plot of the Precipitation Amount (mm) (rain).
(3) Plot of the Highest Gust (knot) (maxgt).
• The user must be able to decide which years to plot. By default it will use the
data from 2015 until 2018.
• The user must be able to decide which plot to draw (i.e, only one of the three,
two of the three, or all three plots). By default the function will create all three
plots.
• The plots must be on a single panel.
• The plots must have meaningful labels and/or titles, and a legend if needed.
• Test your function on the DublinAirport data set

```{r}

plot.WeatherData = function(obj,Begin_year = "2015",End_year = "2018", plot = "X"){
  Duration = subset(obj,obj$year>= Begin_year & obj$year <= End_year)
  Duration$year = as.factor(Duration$year)
  shapes = c("Q1" = 16,"Q2" = 19)
 g1 = ggplot(data = Duration,mapping=aes(x=month))+
   geom_point(aes(y=maxtp,colour=year,shape = "Q1")) +
   geom_point(aes(y=mintp,colour=year,shape = "Q2")) +
   labs(x="Months",y="Variations in Temperature", title= "Monthly Temperature Variations")+
   theme(axis.text.x = element_text(angle = 90,size = 5))+
   scale_shape_manual(name="Temperature",breaks = c("Q1","Q2"),values = shapes,labels = c("Max Temp", "Min Temp"))
  g2 = ggplot(data = Duration,mapping=aes(x=month))+
     geom_point(aes(y=rain,colour=year)) +
   labs(x="Months", y="Variations in Precipitation", title= "Monthly Precipitation Variations")+
    theme(axis.text.x = element_text(angle = 90,size = 5))
  g3 = ggplot(data = Duration,mapping=aes(x=month))+
     geom_point(aes(y=maxgt,colour=year)) +
   labs(x="Months", y="Variations in gust", title= "Monthly gust Variations")+
    theme(axis.text.x = element_text(angle = 90,size = 5))
  if(plot=="1"){
    ggarrange(g1,common.legend = TRUE, legend = "left")
  }                     
  else if(plot=="2"){
    ggarrange(g1,g2,common.legend = TRUE, legend = "left")
  }else if(plot=="3"|plot == "X"){  
    ggarrange(g1,g2,g3,common.legend = TRUE, legend = "left")
  }
}  

plot.WeatherData(obj=DublinAirport)
plot.WeatherData(obj=DublinAirport,Begin_year = "2011",End_year = "2016",plot = 2)
```
